Swarm coordination through perimeter detcetion

### 2017-03-20 paper.tex
First Updates

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Swarm area flooding using field effects}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%
%
%\IEEEcompsocitemizethanks is a special \thanks that produces the bulleted
% lists the Computer Society journals use for "first footnote" author
% affiliations. Use \IEEEcompsocthanksitem which works much like \item
% for each affiliation group. When not in compsoc mode,
% \IEEEcompsocitemizethanks becomes like \thanks and
% \IEEEcompsocthanksitem becomes a line break with idention. This
% facilitates dual compilation, although admittedly the differences in the
% desired content of \author between the different types of papers makes a
% one-size-fits-all approach a daunting prospect. For instance, compsoc 
% journal papers have the author affiliations above the "Manuscript
% received ..."  text while in non-compsoc journals this is reversed. Sigh.

@@

% The paper headers
\markboth{IEEE Transactions on Automatic Control}%
{N.~Eliot: Swarm area flooding using field effects}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

@@

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
\IEEEcompsoctitleabstractindextext{%
\begin{abstract}
%\boldmath
Area flooding is a technique used to fill an enclosed area with agents such that they are distributed as `effectively' as possible throughout an area. An optimum distribution of agents is not necessarily an even distribution. The distribution of the agents is governed by the field effects and the interaction of agents with obstacles (\textit{interaction vectors}).

Filling an area can be applied to tasks that require an unknown environment to be analysed or surveyed. Consider a disaster area following a landslide or a building collapsing following an earthquake; the movement in the land and buildings will produce spaces that are unmapped. The unmapped areas may require investigation to locate people, resources or to create some form of sensor network to analyse the conditions within the area such as creating a heat map or to identify the location toxic gases.

This paper demonstrates two swarm expanding techniques using field effects to perform area filling for the purposes described above.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway. In particular, the Computer Society does
% not want either math or citations to appear in the abstract.

@@

\section{Swarm modelling}
Currently, much swarm research uses field effects as the method of modelling inter-agent interactions~\cite{BAF:06, BAFVM:06, BM:09, APZDAMC:09, GP:02, GP:04, GP:04a, GP:05, GP:11, MYP:09}. The models usually use two field effects to implement the swarming characteristic. These effects are \textit{cohesion}, to draw agents closer, and \textit{repulsion} to prevent agents colliding. Fields are the ranges around an agent that determine the effect other agents have upon its movement~(Figure \ref{methods:FieldEffects}). It is usual for the cohesion field to have a radius $C_b$ which is larger than the repulsion radius $R_b$. When an agent $b'$ moves into the \textit{cohesion field} of an agent $b$ then $b'$ is said to be a neighbour of $b$ and is subject to cohesion. When an agent $b'$ moves into the repulsion field of $b$ then $b$ has a tendency to move away from $b'$, i.e. to be repulsed. When an agent $b$ moves too close to an obstacle, i.e. within the obstacle repulsion range $O_b$, it has a tendency to move away from the obstacle.

\begin{figure}
\begin{center}
\includegraphics[width=6cm]{figures/FieldEffects}
\end{center}
\caption{Agent field effects\label{methods:FieldEffects}}
\end{figure}

Equation (\ref{eq:FlyToCentre1}) identifies the resultant cohesion effect of all the neighbours of $b$. $nbr(.)$ returns a set of all the agents in the swarm that are a neighbour of $b$. A neighbour is any agent that fails within the cohesion field.

\begin{equation}\label{eq:FlyToCentre1}
v_{c}(b) = \frac{\mathlarger{\sum_{b' \in nbr(b)}}{bb'}}{|nbr(b)|}
\end{equation}‎

Equation (\ref{eq:Repulsion1}) identifies the resultant repulsion effect of all the neighbours of $b$ where $rep(.)$ returns a set of all agents that are within the repulsion field.

\begin{equation}
\label{eq:Repulsion1}
v_{r}(b) =‎ -
\frac{1}{|R(b)|}
\left(
\mathlarger{\mathlarger{\sum_{b' \in rep(b)}}}
{\left( 1-\frac{|bb'|}{R_b} \right)}
{bb'}
\right)
\end{equation}‎

Using the cohesion and repulsion vectors generated by the relationship of $b$ to its neighbour a resultant vector can be calculated. This vector creates an agent characteristic that can be used as a metric. Summing the vectors creates a resultant vector with a magnitude that affects the agent. Summing the vectors also provides an indication of the direction an agent will move based on the relationship. This is the \textit{inter-agent vector}.

Equation (\ref{eq:InterAgentMovement1}) identifies the \textit{inter-agent vector} for agent $b$ with respect to its neighbours.

\begin{equation}\label{eq:InterAgentMovement1}
v(b) = v_{c}(b) + v_{r}(b)
\end{equation}‎

If the system is applying a weighted model then equation~\ref{eq:InterAgentMovement1} becomes:

\begin{equation}\label{eq:InterAgentMovement2}
v(b) = k_cv_{c}(b) + k_rv_{r}(b)
\end{equation}‎

To implement the boundary of an area a third component needs to be added to the model in the form of obstacles.

Obstacles, like agents, can be represented as a point in the system. As an agent moves an obstacles may enter its \textit{obstacle repulsion field} causing the agent move away.

In this thesis agents are modelled with a fixed obstacle repulsion distance $O_b$ where a repulsion vector is applied. The repulsion is then a vector of magnitude $O_b$. If more than one obstacle is within the field effect agent the total repulsion vector is the sum of the repulsion vectors due to each obstacle~(Figure~\ref{methods:Obstacle1}). The result is normalised and scaled such that the magnitude is the same as the field distance~$O_b$.

\begin{figure}
\begin{center}
\includegraphics[width=6cm]{figures/Obstacle1}
\end{center}
\caption{Obstacle repulsion \label{methods:Obstacle1}}
\end{figure}

%% $O$ is a set of $N$ obstacles.
%% 
%% \begin{center} \label{eq:Obstacle1}
%% \begin{equation}‎
%% O \buildrel \Delta \over =‎ \{o_1\ldots o_N\}
%% \end{equation}‎
%% \end{center}

%% $o' \in O:|bo'| < O_b$ the set of obstacles that are within range of agent $b$, where $O$ is the set of obstacles.

%% \begin{center} \label{eq:Obstacle4}
%% \begin{equation}‎
%% R(b) \buildrel \Delta \over =‎ \{o' \in O:|bo'| < O_b\}
%% \end{equation}‎
%% \end{center}

Figure~\ref{eq:Obstacle2} shows the resultant repulsion vector $v_o(b)$ for an agent. $\{o\ \in O:|bo| <= O_b\}$ is the set of obstacles that are within range of agent $b$. $O$ is the set of obstacles. The obstacles are identified using the distance between an agent and an obstacle $|bo|$ and comparing the result to the fixed obstacle repulsion range~$O_b$. The result is calculated by scaling the normalised sum of the normalised vectors $(ob)\string^$ by $O_b$.

\begin{center} \label{eq:Obstacle2}
\begin{equation}‎
v_o(b) =‎ O_b\Bigg(\mathlarger{\mathlarger{\mathlarger{\sum}}}_{o \in O:|ob| <= O_b}(ob)\string^\Bigg)\string^
\end{equation}‎
\end{center}

%% \begin{center} \label{eq:Obstacle3}
%% \begin{equation}‎
%% v_o(b) =‎ |l_o(b)|O_b
%% \end{equation}‎
%% \end{center}as shown in equation~\ref{eq:InterAgentMovement3}.

The resultant swarm model is shown in~equation~\ref{eq:InterAgentMovement3}:

\begin{equation}\label{eq:InterAgentMovement3}
v(b) = k_cv_{c}(b) + k_rv_{r}(b) + k_ov_{o}(b)
\end{equation}‎

\section{Area flooding}\label{chapter:flooding}
The concept of using a swarm to provide coverage over unknown areas is a current area of research. Alvissalim et al. (2012) discuss the application of commercially available drones to provide a communication infrastructure across an unknown disaster areas~\cite{AZHMJJM:12}. Scheutz and Bauer (2006) use both cohesion and repulsion as a mechanism to create coverage of an area that requires protection in an adversarial environment~\cite{SB:06}. In their paper they do not use cohesion to ensure the swarm remains a cohesive unit, rather they have all the agents using a \textit{destination vector} to cause agents to converge on a target and the repulsion vector prevents collisions. The cohesive effect is not required due to all agents having a common goal with no requirement for the agents to remain in close proximity during the terrain traversal. Ramaithitima et al. in 2015 and 2016, working as part of Vijay Kumar's research group at the University of Pennsylvania, use the idea of filling an area through a deployment strategy that requires the swarm to identify the placement of agents in a given space. The agents are then `shuffled' into an identified position. The technique applied in 2015~\cite{RWBK:15} is based upon coverage path planning and involves adding agents when required and does not require a global positioning reference. In their 2016 paper~\cite{RWBK:16} they use Voronoi graphs to achieve the same effect. Yang et al. (2015)~\cite{YDH:15} also use Voronoi graphs to achieve their swarm coverage goal. The use of Voronoi graphs requires inter-agent communications. Schroeder and M. Kumar (2016) use the bio-inspired swarming technique of creating chemical trails (pheromones) combined with a foraging approach which they refer to as a `food foraging' technique~\cite{SK:16}.

The techniques used in this paper are similar to Schroeder and M. Kumar in that the agents do not require a global positioning system and differ from Alvissalim et al., Scheutz and Bauer, and Ramaithitima et al. in that the space filling is achieved with a fixed sized swarm.

The expansion fill is implemented using two techniques. The main principle behind both techniques is to increase the repulsion field effect of the agents over time to increase the area coverage of the agents. The expansion fill can use both the cohesion and repulsion field effects in a similar manner to a static swarm in free space. This technique ensures all the agents remain in close proximity and as far as possible the agents form a single entity. Alternatively the expansion fill can be implemented using only a repulsion field effect. Cohesion can be eliminated in an enclosed space as the agents have a limited range of movement (bounded by a perimeter). When agents move to a repulsion boundary, either an area perimeter or an obstacle, they are repelled. As the space is finite the repulsion `pushes' the agents back into the swarm countering the expansion that is induced by the \textit{interaction vectors}. If the repulsion field effect creates a swarm that reaches all the boundaries the repulsion imposes a compression effect as the \textit{interaction-vector} `pushes' the agents into the boundaries.

\section{Field effect modification with cohesion and repulsion}
The concept behind the field effect expansion is to increase both the repulsion and cohesion fields over a period of time. This increase in the field effects makes the agents increase the distance between each other expanding the swarm as a whole. The expansion increases until, due to boundary compression, the swarm is unable to expand further. In an extreme case the expansion will result in a set of field effects that create a mesh based swarm structure rather than the desired hexagonal structure that the field effects should create. Identifying either the change in modality or the inability to expand, which can be identified by the \textit{inter-agent vector magnitude} increase or a reduction in the inter-agent distance stability indicates area saturation has occured.

To test this hypothesis a swarm is modelled in the simulator. The model consists of an obstacle-based enclosed space and a swarm consisting of 60 agents. The experiments parameters for the simulation are shown in~Table~\ref{tab:FillParameters}

\begin{table}
\begin{center}
\begin{tabular}{| p{1.5cm} | p{1.5cm} | p{4.0cm} |}
\hline
\bf Parameter & \bf Value  & \bf Description \\ \hline
$k_c$         & 5          & weight adjuster for cohesion bias\\ \hline
$k_r$         & 60         & weight adjuster for repulsion bias\\ \hline
Sample rate   & 100ms      & proximity sensor rate\\ \hline
Speed         & 20 units/s & agent speed\\ \hline
\end{tabular}\caption{Swarm parameters} \label{tab:FillParameters}
\end{center}
\end{table}

The field effects are incremented in turn, neighbour range followed by repulsion range. After each repulsion field change the swarm is allowed to redistribute itself and stabilise. Table~\ref{tab:BaselineConcaveReduction} shows the field settings that are used for the simulation. The field effect are selected so as to ensure the swarm parameters have the potential to create a hexagonal swarm.

\begin{table}
\begin{center}
\begin{tabular}{| p{1.5cm} | p{0.3cm} | p{0.3cm} | p{0.3cm} | p{0.3cm} | p{0.3cm} | p{0.3cm} | p{0.3cm} | p{0.5cm} |}
\hline
\bf Weight \bf component & \bf 1 & \bf 2 & \bf 3 & \bf 4 & \bf 5 & \bf 6 & \bf 7 & \\ \hline
Repulsion Boundary & 50 & -  & 60 & -  & 70 & -  & 80 & units\\  \hline
Neighbour Distance & 60 & 70 & -  & 80 & -  & 90 & -  & units\\  \hline
\end{tabular}\caption{field effect expansion sequence} \label{tab:FillSequence}
\end{center}
\end{table}

Figures~\ref{emerge:Expand1}, \ref{emerge:Expand2}, \ref{emerge:Expand3}, \ref{emerge:Expand4}, \ref{emerge:Expand5} and \ref{emerge:Expand6} show the stages that the swarm progresses through during the simulation. Figure~\ref{emerge:Expand1} shows the initial deployment of the 60 agents within the enclosed environment. Figure~\ref{emerge:Expand6} shows the stage at which the experiment terminates due to area saturation.

\begin{figure}
\begin{center}
\includegraphics[width=5.5cm]{figures/EXPAND1}
\end{center}
\caption{Stage 1\label{emerge:Expand1}}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=5.5cm]{figures/EXPAND2}
\end{center}
\caption{Stage 2\label{emerge:Expand2}}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=5.5cm]{figures/EXPAND3}
\end{center}
\caption{Stage 3\label{emerge:Expand3}}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=5.5cm]{figures/EXPAND4}
\end{center}
\caption{Stage 4\label{emerge:Expand4}}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=5.5cm]{figures/EXPAND5}
\end{center}
\caption{Stage 5\label{emerge:Expand5}}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=5.5cm]{figures/EXPAND6}
\end{center}
\caption{Stage 6\label{emerge:Expand6}}
\end{figure}

%% \begin{figure}
%% \centering
%% \subfigure[Stage 1]{
%% 	 \includegraphics[width=5.5cm]{figures/EXPAND1}
%% 	 \label{emerge:Expand1}
%% }
%% \subfigure[Stage 2]{
%% 	 \includegraphics[width=5.5cm]{figures/EXPAND2}
%% 	 \label{emerge:Expand2}
%% }
%% \subfigure[Stage 3]{
%% 	 \includegraphics[width=5.5cm]{figures/EXPAND3}
%% 	 \label{emerge:Expand3}
%% }
%% \subfigure[Stage 4]{
%% 	 \includegraphics[width=5.5cm]{figures/EXPAND4}
%% 	 \label{emerge:Expand4}
%% }
%% \subfigure[Stage 5]{
%% 	 \includegraphics[width=5.5cm]{figures/EXPAND5}
%% 	 \label{emerge:Expand5}
%% }
%% \subfigure[Stage 6]{
%% 	 \includegraphics[width=5.5cm]{figures/EXPAND6}
%% 	 \label{emerge:Expand6}
%% }
%% \caption{Space filling via field effect expansion}
%% \label{methods:ExpansionFill}
%% \end{figure}

After the initial deployment the system is allowed to settle~Figure~(\ref{emerge:Expand2}) following the settling period the parameters are increased and the swarm settles again. Figure~\ref{emerge:Expand3}~shows how the change in parameters can cause the system to become unstable, the top right section of the swarm has became multi-modal yet there is still a section of the swarm that has not expanded to the boundary perimeter. As the swarm stabilises this compressed section of the swarm `pushes' through the swarm and the swarm increases in volume filling more of the space~(Figure~\ref{emerge:Expand4}). This process continues until changes in the parameters are unable to increase the distribution of the agents and the distances remain almost constant. Increasing the field effects further causes the swarm to become multi-modal as shown in~Figures~\ref{emerge:Expand5} and~\ref{emerge:Expand6}.

These effects can be identified through a combination of the \textit{inter-agent magnitude metric} and the distance metric.

\subsection{Magnitude analysis}
Figure~\ref{emerge:FIELDFILL-MAG} shows the resultant magnitude for the swarm for the entire simulation. Between 100 seconds and 105 seconds there is a significant change in the magnitude where the value becomes negative. This indicates there is a compression effect on the swarm and it is unable to expand the inter-agent distances.

\begin{figure}
\begin{center}
\includegraphics[width=8.5cm]{figures/FIELDFILL-MAG}
\end{center}
\caption{Magnitude metric 0-120 seconds\label{emerge:FIELDFILL-MAG}}
\end{figure}

Figure~\ref{emerge:FIELDFILL-MAG-1} shows the initial deployment status of the swarm and the first repulsion increase at 25 seconds. When the repulsion field is increased, the average \textit{inter-agent magnitude} falls and the variation increases. The agents redistribute themselves which takes until approximately 65 seconds and a new baseline is established for the field effects.

>>>>

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Swarm coordination through perimeter detcetion}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%
%
%\IEEEcompsocitemizethanks is a special \thanks that produces the bulleted
% lists the Computer Society journals use for "first footnote" author
% affiliations. Use \IEEEcompsocthanksitem which works much like \item
% for each affiliation group. When not in compsoc mode,
% \IEEEcompsocitemizethanks becomes like \thanks and
% \IEEEcompsocthanksitem becomes a line break with idention. This
% facilitates dual compilation, although admittedly the differences in the
% desired content of \author between the different types of papers makes a
% one-size-fits-all approach a daunting prospect. For instance, compsoc 
% journal papers have the author affiliations above the "Manuscript
% received ..."  text while in non-compsoc journals this is reversed. Sigh.

@@

% The paper headers
\markboth{IEEE Transactions on Automatic Control}%
{N.~Eliot: Swarm coordination through perimeter detection}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

@@

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
\IEEEcompsoctitleabstractindextext{%
\begin{abstract}
%\boldmath
This chapter identifies the effect a \textit{destination vector} has upon the motion of a swarm. Three coordination techniques are applied to a swarm and metrics are used to identify the effects. The three coordination techniques are: full perimeter detection, where only edge based agents will have a \textit{destination vector} applied. Partial perimeter detection, where a subset of the perimeter agents are detected and finally all agents in the swarm having a \textit{destination vector} applied. For all the experiments the base parameters of the swarm are fixed ensuring that the only variable will be the \textit{destination vector} that is applied by the coordinating agents. This process allows the impact of the algorithms to be isolated and compared. Section~\ref{sec:AlternateBias1}~discusses the potential variations to the base parameters and the effects this has on controlling the impact of the weightings.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway. In particular, the Computer Society does
% not want either math or citations to appear in the abstract.

@@

%% \section{Introduction}
\section{Baseline specification}
To allow an analysis of each algorithm's effect on a swarm a baseline measurement of `background' variance (jitter) is identified. The baseline measurement provides a comparative data set for the experiments. The baseline data set is for a static swarm (no destination vector) with the same internal parameters.

\textit{Assumption 1:} The swarm used for the experiments consists of 200 agents randomly distributed~(Figure~\ref{fig:SampleSwarm200-1}).\\ 
\textit{Assumption 2:} The field effects and bias will be set as shown in~Figure~\ref{tab:DynamicPhysics1} with the weight adjuster $k_d = 0$ for the baseline.\\
\textit{Assumption 3:} All destination based experiments will have the weight adjuster set to $k_d = 100$.

\begin{figure}
\begin{center}
\includegraphics[width=14cm]{figures/SampleSwarm200-1}
\end{center}
\caption[Sample swarm 200 agents initial state]{Sample swarm 200 agents initial state (\textit{screen shot from simulator})\label{fig:SampleSwarm200-1}}
\end{figure}

\begin{table}
\begin{center}
\begin{tabular}{| p{2.5cm} | c | p{7cm} |}
\hline
\bf Weight \bf Component & \bf Hexagonal & \bf Description \\ \hline
Sample Rate & 100 & ms - Unit sampling interval\\  \hline
$k_c$ & 5 & weight adjuster for cohesion vector\\  \hline
$k_r$ & 15 & weight adjuster for repulsion  vector\\  \hline
$k_d$ & 0/100 & weight adjuster for destination vector 0 for static baseline 100 for destination based\\  \hline
Neighbour Distance & 70 & units\\  \hline
Agent Boundary & 80 & units\\  \hline
Speed & 20 & units/s\\  \hline
\end{tabular}\caption{Swarm Weighted Model} \label{tab:DynamicPhysics1}
\end{center}
\end{table}

The results for the baseline experimental swarm are shown in~Figures~\ref{coord:BaselineMagnitude1} and~\ref{coord:BaselineDistance1}. The metrics used are the distance \cite{NIM:09} and resultant inter-agent magnitudes (\textit{inter-agent vector}).

\begin{figure}
\begin{center}
\includegraphics[width=14cm]{figures/BaselineMagnitude1}
\end{center}
\caption{Baseline internal movement - magnitude\label{coord:BaselineMagnitude1}}
\end{figure}

Figure~\ref{coord:BaselineMagnitude1} shows the magnitude between $t_0$ and approximately $t_{48}$ dominated by repulsion ($|k_rv_r| > |k_cv_c|$). This indicates that the swarm is undergoing a rapid expansion as the bias within the swarm shows a high level of repulsion. Between approximately $t_{48}$ and $t_{100}$ the magnitude is positive but rising which indicates the swarm is cohesive and will remain a cohesive unit but it is still expanding. At approximately $t_{110}$ the swarm has completed it's expansion and has reached a settled state with the `background' movement maintaining the swarm's structure with $|k_rv_r| < |k_cv_c|$. The residual `jitter' is at its minimum for the configuration and the agent distribution is at its optimum for the cohesion and repulsion settings. The resultant $|k_rv_r| < |k_cv_c|$ is due to the cohesion field being larger than the repulsion field which prevents the swarm `breaking up'.

\begin{figure}
\begin{center}
\includegraphics[width=14cm]{figures/BaselineDistance1}
\end{center}
\caption{Baseline internal movement - distance\label{coord:BaselineDistance1}}
\end{figure}

Figure~\ref{coord:BaselineDistance1} shows the internal movement and the deviation from the mean for the swarm based on the inter-agent distances. There is no \textit{destination vector} applied so all internal movement is being generated by the cohesion and repulsion vectors. The swarm is initially in a chaotic state but once the agents have expanded a stable hexagonal formation evolves. Figure~\ref{coord:BaselineDistance1} shows the first 30s of the simulation which is sufficient time for the swarm to `settle' and transform to its most stable state. Due to the metric only showing inter-agent distances it is not possible to determine if the swarm will remain cohesive.

The centroid path for the baseline swarm is shown in~FIgure~\ref{coord:SwarmPathBaseline1}. Due to there being no \textit{destination vector} for the swarm the centroid changes positions based on the inter-agent positional changes caused by the \textit{interaction vectors}. These changes are the result of each agents attempting to attain a state of equilibrium within their `clusters'.

\begin{figure}
\begin{center}
\includegraphics[width=14cm]{figures/SwarmPathBaseline}
\end{center}
\caption{Baseline swarm path\label{coord:SwarmPathBaseline1}}
\end{figure}

These baseline measurements are used to identify the changes that occur in the structure of a swarm when changes are made to the weightings of the field effects, the number of coordination agents and the \textit{destination vector} weighting.

\section{Destination vector application}
Many mechanisms can be used to influence a swarm to travel in a specific direction. These mechanisms generally involve the use of sensors such as magnetometers or GPS's to achieve a directional coordinate \cite{SG:15,TG:12}.

This thesis aims to produce algorithms that will reduce the number of active sensors that are needed for the directional coordination of a swarm. When reducing the number of active sensors the algorithms must maintain sufficient coordination of agents to influence the swarm's direction and therefore create an overall \textit{destination vector} to a swarms movement. Reducing the number of active sensors will reduce the net energy usage of the sensors and potentially increase the sensors viable lifetime. Another aspect of the algorithms is that they should minimise the internal disturbance (jitter) created by the \textit{destination vector}.

There are two distinct approaches to configuring agents: all the agents are configured identically (homogeneous swarm) or the agents are in a selection of different configurations (heterogeneous swarm)~\cite{BS:13}.

The following assumptions are made in this thesis:

\textit{Assumption 1:} All agents are identical (homogeneous) but are able select subsystems as necessary and adopt different roles depending upon the algorithm being used to coordinate the swarm. \\
\textit{Assumption 2:} As the swarm progresses towards a destination the role of an agent can change such that all the agents are identical in construct but they can enable and disable sensors as necessary. This will allow agents within the swarm to self promote to the coordinator role should a set of conditions arise.\\
\textit{Assumption 3:} All agent in the swarm are autonomous and each agent has its own power supply.

Reducing the usage of sensors will reduce the net power consumption of the swarm and increase the time that an agent can be a part of a swarm. If the purpose of the swarm is reconnaissance then the reduction in energy usage may allow a longer traversal time, or allow the swarm to travel further. As agents in the swarm will use there resources at different rates some agents may be lost due to resource exhaustion. The algorithms should therefore be resilient to this.

\section{Swarm destination vector}
The following assumptions are made in this thesis:

\textit{Assumption 1:} The direction of the swarm is based upon having a fixed goal that is known to all the agents (Section \ref{sec:Direction1}). \\
\textit{Assumption 2:} The goal is migrated towards based on a simulated GPS signal and the bias $k_d$ is applied to the \textit{destination vector} to influence the movement of the agents.

The swarm's destination is established by the coordinator agents influencing the non-coordinator agents. This influence is applied through the coordinators proximity to non-coordinator agents (cohesion). Coordinator agents are identified by a set of conditions. The conditions are the activators for the coordinator identification algorithms.

\section{Identifying the coordinator role}
For an agent to change its role from a standard agent to a coordinator there needs to an algorithm that identifies a set of conditions to trigger the change. This same algorithm also triggers an agent to revert to being a standard agent. The coordinator selection algorithm is separate from the agent coordination algorithms discussed in chapter~\ref{chapter:methods}.

In this thesis three algorithms are presented and compared. In the first algorithm, all agents have the role of a coordinator permanently. This will be referred to as the all-agent algorithm. The other two mechanisms employ a selection criteria to identify if the simulated GPS sensor should be enabled. These two algorithms are: full-perimeter detection and the partial perimeter detection. Partial perimeter detection will be referred to as the basic-count algorithm.

There are many other ways in which the coordinator role could be established such as randomly enabling a GPS for a set period of time or oscillating the role on and off but they are beyond the scope of this thesis.

\section{Monolithic swarm - (all-agent)} 
In a monolithic swarm \cite{BS:13} the propagation towards the goal is achieved by all the agents using a GPS signal to give each agents' movement a \textit{direction vector} towards a required goal~(Equation~\ref{eq:BotDirection1}). There is no selection criteria and no computational overhead for the agents.

Figure~\ref{reduced:Monolithic1} shows a screen shot from the swarm simulator with all agents GPS's enabled. This can be seen as the agents highlighted with a ring around them. The small `tail' on each agent is the \textit{movement vector}.

\begin{figure}
\begin{center}
\includegraphics[width=7cm]{figures/Monolithic2}
\end{center}
\caption[Monolithic agents]{Monolithic agents (Circles indicate all agents are using GPS, \textit{screen shot from simulator}) \label{reduced:Monolithic1}}
\end{figure}

\subsection{Baseline and effect of no perimeter detection\label{section:compareBaselineNoPerimeter}}
To identify the effect of the all-agents algorithm the simulation is executed using the same parameters as the baseline swarm with the added \textit{destination vector} on all the agents. The effect of introducing this bias on the swarms internal characteristics is shown in~Figures~\ref{reduced:BaselineAllAgentsMag1} and~\ref{reduced:BaselineAllAgentsDist1}.
%BASELINE-ALLAGENTS-MAG.py
\begin{figure}
\begin{center}
\includegraphics[width=14cm]{figures/BaselineAllAgentsMag1}
\end{center}
\caption{Baseline/All agents comparison \textit{inter-agent vector} magnitude\label{reduced:BaselineAllAgentsMag1}}
\end{figure}

Figure~\ref{reduced:BaselineAllAgentsMag1} shows the effect on the \textit{inter-agent vector} magnitude of the swarm: the average resultant magnitude is lower than the baseline. This is due to there being a greater distribution of \textit{inter-agent vector} magnitudes from the disturbance caused by the \textit{destination vector} on all of the agents and the swarm is therefore less cohesive. The graph also shows that there is a higher variation from the mean; again this is due to the change in the bias of all the agents. The agents are moving towards a goal and are therefore not moving to an equilibrium distribution. Figure~\ref{reduced:BaselineAllAgentsMag1}~also shows that the \textit{direction vector} increases the time it takes for the swarm to distribute the agents and although the deviation diminishes the swarm cannot overcome the \textit{direction vector's} effect and the swarm appears chaotic.
%BASELINE-ALLAGENTS-DIST.py
\begin{figure}
\begin{center}
\includegraphics[width=14cm]{figures/BaselineAllAgentsDist1}
\end{center}
\caption{Baseline/All agents comparison (distance)\label{reduced:BaselineAllAgentsDist1}}
\end{figure}

Figure~\ref{reduced:BaselineAllAgentsDist1} shows the effect on the inter-agent distances. The average distance is lower than the baseline. This effect is caused by the \textit{direction vector} reducing the effect of the repulsion field. The agents therefore compress slightly due to the cohesion.

A curious effect of having all the GPS sensors enabled is that as the swarm approaches the destination there is an increased compression effect and the internal disturbance increases. This is the result of the agents `pulling' in a cone effect. The agents paths converge on the destination~Figure~\ref{reduced:ConicalPath}. This effect is caused by the number of agents being effected by a \textit{direction vector}. This effect is less identifiable if a swarm's target is at a greater distance.

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{figures/ConicalPath}
\end{center}
\caption{Conical destination trajectories\label{reduced:ConicalPath}}
\end{figure}

This effect can be seen in both the magnitudes shown in~Figure~\ref{reduced:BaselineAllAgentsMag2} and in the distances shown in~Figure~\ref{reduced:BaselineAllAgentsDist2} which both show the averages initially decreasing but then increasing again as the swarm approaches the destination. This is accompanied by a greater standard deviation caused by the \textit{direction vector} affecting the priority of the swarm as agents are drawn to the destination reducing the effect of the \textit{inter-agent vectors} creating stable structures.
%BASELINE-ALLAGENTS-MAG.py
\begin{figure}
\begin{center}
\includegraphics[width=14cm]{figures/BaselineAllAgentsMag2}
\end{center}
\caption[Baseline/All agents magnitude comparison]{Baseline/All agents \textit{inter-agent vector} magnitude comparison (60 seconds)\label{reduced:BaselineAllAgentsMag2}}
\end{figure}
%BASELINE-ALLAGENTS-DIST.py
\begin{figure}
\begin{center}
\includegraphics[width=14cm]{figures/BaselineAllAgentsDist2}
\end{center}
\caption{Baseline/All agents distance comparison (60 seconds)\label{reduced:BaselineAllAgentsDist2}}
\end{figure}

\section{Simple multifaceted swarm (basic-count)} 
In a simple multifaceted swarm the propagation towards the goal is achieved by generating a \textit{destination vector} for an agent when the agent conforms to a simple counting rule. This \textit{destination vector} is obtained by using the agent's GPS to identify its current coordinate and generating a vector from that coordinate to a specific destination point. This rule creates a subset of agents that apply a local \textit{direction vector} to their movement. This has the effect of creating a directional effect on the whole swarm. The aggregate \textit{direction vector} magnitude of the swarm is less than that produced when all agents apply a \textit{direction vector}. This reduction is further diluted by agents without the \textit{direction vector} calculating a \textit{movement vector} to an equilibrium position~(Equation~\ref{eq:BotDirection1}). This change in the aggregate \textit{direction vector} will impact on a swarm's `jitter'. This is discussed in more detail in~Section~\ref{sec:AlternateBias1}. With there only being a limited number of peripheral agents invoking their GPS signal there will also be a net reduction in the energy usage of the swarm~(Figure~\ref{reduced:EnergyReductionGPS}).

A swarm with the appropriate field effects will propagate towards hexagons made from a central agent and six neighbours~(Figure~\ref{fig:StableForms}). A swarm can be influenced by its surroundings, which involves the interaction of other agents (as neighbours) and obstacles which can cause a compression of the agents. Compression can also occur during the stabilisation phase of a swarm. These situations can cause an agent to have less than 6 neighbours but still be surrounded. A general rule therefore to detect a rudimentary boundary of a hexagonal swarm is to calculate the number of neighbours an agent has. If the total is less than five ($nbr(b) < 5$) then there is a high probability that the agent is either in a void within the swarm or on an external perimeter. If an agent is on a perimeter or a void the it should enable its GPS and become a coordinator and provide a directional bias to the swarm via its \textit{direction vector}. This coordinator role affects the immediate neighbours \textit{movement vector} through cohesion/repulsion~(Figure~\ref{reduced:SimpleMulti12}). This simple detection algorithm allows a directional bias to be induced into a swarm with minimal computational impact on the coordinator agents.

\subsection{Simple multifaceted algorithm}
The mechanism to enable a coordinator agent is to count the number of neighbours that an agent has~(Algorithm~\ref{algo:neighbourDetect}). This process is already part of the cohesion and repulsion calculations.
